### 2023-09-23 chapters/background.tex
Work on variational autoencoders

While the papers in this thesis were written to be self-contained, page limits and the need to maintain a specific focus, forced us to limit the scope of technical background each paper provides. In this chapter, we present a more in-depth overview of the technical background relevant to the papers in this thesis. 
We start by quantifying the concept of uncertainty, and how it relates to information and probability. We then introduce the task of out-of-distribution detection and a few ways to tackle it. 
Then, we present variational autoencoders, and how they can be used to learn representations of speech. Finally, we introduce the concept of self-supervised learning and highlight important differences to variational autoencoders.

@@

In society, the concept of uncertainty goes by many names, and its meaning can vary depending on the specific context. However, across most quantitative scientific fields, a consensus definition appears to align with the following \cite{hubbard_how_2014}:
%
\begin{center}
    \textit{Uncertainty is the lack of certainty; a state of limited knowledge where it is impossible to exactly describe the existing state, a future outcome, or more than one possible outcome.}
\end{center}
%
Although such a purely lexical definition of uncertainty might prompt a philosophical inquiry, it highlights an important connection between uncertainty, information and probability. When information is limited, we can describe the state of the world only with uncertainty; we are simply not certain. In that sense, uncertainty is a word we use to refer to the information that is missing, whether we know what is missing or not. 
A natural way to describe an uncertain world state is to use probabilities. To our luck, information theory provides exactly that; a mathematically rigorous method for quantifying information through the language of probability theory. In the following, we shall see that, in this context, uncertainty can be understood as the information entropy of a random variable \cite{mackay_information_2003}.

@@

The \emph{information entropy of a discrete random variable} $X$ is defined as the \emph{expected information content} of an outcome of $X$,
%
\begin{equation} \label{eq:information-entropy}
    H(X) = \mathbb{E}_X\left[I_X(x)\right] = - \sum_{x\in X} p_X(x) \log p_X(x) \enspace .
\end{equation}
%
This means that $H(X)$ measures the amount of information we can expect to gain from observing an outcome of $X$, when we know only its distribution $p_X(x)$. If all possible outcomes are equally likely, then $H(X)$ is maximized, and if only one outcome is possible, then $H(X)$ is minimized.

@@

To generalize the concept of information entropy to continuous random variables, Shannon originally replaced the sum over the probability mass function in \eqref{eq:information-entropy} with an integral over the probability density function, as suggested by the definition $H(X) = \mathbb{E}_X\left[I_X(x)\right]$. This approach leads to the definition of the \emph{information entropy of a continuous random variable} $X$ as, 
%
\begin{equation} \label{eq:differential-entropy}
    H(X) = - \int_{\mathcal{X}} p_X(x) \log p_X(x) \, dx \enspace ,
\end{equation}
%
which is called the \emph{differential entropy} \cite{shannon_mathematical_1948}. Here, $\mathcal{X}$ is the support of $X$.

However, the differential entropy does not have several of the desirable attributes of the discrete version; it can be negative, it is not invariant under a change of variables, and it is not dimensionally correct%
\footnote{\label{fn:dimensional-analysis-of-differential-entropy}
    Information entropy has the same dimensionality as information content which is typically measured in units of bits, but depends on the base of the logarithm. 
    However, the bit is not a base unit of the International System of Units (SI), nor is it an official derived unit.
    A bit is quite simply a number, specifically 0 or 1. If something is said to have a size of 4 bits, it means that it can be described with 4 binary digits, i.e. 4 numbers, each either 0 or 1.

@@

\section{Out-of-distribution detection} \label{sec:out-of-distribution-detection}
% 
% OUTLINE: C.f. https://openaccess.thecvf.com/content/CVPR2023W/VAND/papers/Graham_Denoising_Diffusion_Models_for_Out-of-Distribution_Detection_CVPRW_2023_paper.pdf
% - Formalize the problem. https://arxiv.org/pdf/2302.10326.pdf
%   - Assume we have a dataset $\mathcal{D}_{\text{train}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{train}}$ or not. Given a test point $\xb$, the model outputs a score $s(\xb)$ that indicates whether $\xb$ is in-distribution or out-of-distribution. 
%   - In the context of neural networks, this problem dates back to at least 1994 \cite{bishop_novelty_1994}. 
%
% - Categorisation into supervised and unsupervised. 
%   - Supervised: Whether p(y|x) should be trusted for a given x. 
%   - Unsupervised: Whether x should be trusted and used at all. 
%
In this section we will introduce out-of-distribution (OOD) detection, relate it to the concept of uncertainty and to the previous section on information entropy. 
To supplement the short related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, we will then provide a concise review of the literature on out-of-distribution detection. A

\subsection{An overview of the task}
%
OOD detection is the task of identifying test data that is unlikely to originate from the distribution of the training data and, in the context of neural networks, dates back several decades \cite{bishop_novelty_1994, chang_figure_1993}.
In the general case, we assume that we have a domain of in-distribution data, $\mathcal{D}_{\text{in}}$, and we would like to build a model that can be used to assess whether a test data point $\xb$ originates from that domain or not. 
Since such OOD data is unknown to the model, it is, by definition, a source of epistemic uncertainty. This makes OOD detection closely related to uncertainty quantification. Only in cases where strong prior knowledge of the data is incorporated into a model, can we expect it to be able to make reliable predictions on OOD data.

% In the literature, the task is also referred to as open set recognition \cite{bendale_open_2016, cardoso_weightless_2017, panaredabusto_open_2017}, outlier detection \cite{xu_deep_2014, mor_confidence_2018}, or selective prediction \cite{geifman_selective_2017} . 
OOD detection bears many similarities with anomaly detection, novelty detection, open set recognition, and outlier detection \cite{yang_generalized_2022}. 
Some of these differences are subtle, and the terminology is not always used consistently in the literature. To provide some clarity, we will briefly outline the taxonomy of \textcite{yang_generalized_2022}: 
Outlier detection most clearly differs from the other tasks by directly processesing all observations and aiming to select outliers from a single contaminated dataset. 
The remaining tasks differ in whether they detect both covariate and semantic shift and require the simultaneous classification of in-distribution classes.%, and whether the in-distribution data contains more than one class. 
Anomaly detection deals with multiclass data and detects both covariate and semantic shift without requiring simultaneous classification of in-distribution classes whereas novelty detection, open set recognition and OOD detection are usually concerned with semantic shifts. 
Although a vague difference, novelty detection usually defined in terms of a class of normal data, while OOD detection centers around the distribution of the training data. Different from novelty detection, OOD detection methods sometimes draw on examples of OOD data or require the simultaneous classification of in-distribution classes, although this is not always the case. 
OOD detection benchmarks almost always take OOD data to be from external datasets different from the training dataset which distinguishes it from open set recognition that usually benchmarks on a single dataset split into ID and OOD data. 
% Another closely related field is that of network-based uncertainty estimation \cite{neal_bayesian_1995, hernandez-lobato_probabilistic_2015, gawlikowski_survey_2023}. %[14, 10, 36, 20]. 
Although we share the sentiment of \textcite{yang_generalized_2022}, who propose to unify the tasks as ``generalized out-of-distribution detection'', this thesis will follow the nomenclature used in the works most related to ours and refer to the task as out-of-distribution detection.

Recently, several approaches for deep neural networks have been developed that formally address the rejection of samples $\xb\notin\mathcal{D}_{\text{in}}$. 
One way to categorize the different approaches is based on whether the underlying model is a classifier that estimates a conditional probability distribution $p(y|\xb)$ over some target variable given the input, or a model that learns a probability density $p(\xb)$ over the input itself. We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \cite{graham_denoising_2023,liu_unsupervised_2023a}. It is important to note that this distinction relates only to the availability of some target value, $y$ - not whether OOD data is available for supervision\footnote{
    In our overview, we distinguish between supervised and unsupervised out-of-distribution by letting methods be classified as supervised that use any kind of target value $y$, whether it relates to the original task, available OOD data, or both. This is the same distinction made by \textcite{graham_denoising_2023} and \textcite{liu_unsupervised_2023a}. 
    However, it is important to note that in other works, the distinction is made based on whether the model is trained on OOD data or not \cite{hendrycks_baseline_2017,liu_energy-based_2020}. This difference in nomenclature is currently unresolved in the literature. 
}. 
Supervised OOD detection tries to assess whether the model's prediction $\hat{y}$ via $p(y|\xb)$ should be trusted for a given input $\xb$, whereas unsupervised OOD detection judges whether the input $\xb$ should be trusted and used at all.

Central to any OOD detection method is the ability to assign a score $s(\xb) \in \mathbb{R}$ to a given input $\xb$ that indicates the degree to which the input is likely to be OOD. After defining a score, we typically use a validation set to tune a threshold $\tau$ such that $\xb$ is considered OOD if $s(\xb) > \tau$. The threshold is typically chosen such that performance on the validation set is above some standard, for instance by imposing constraints on the recall and precision. Many works also evaluate the performance of OOD detection methods using the area under the receiver operating characteristic (AUROC) curve which does not require a threshold to be chosen.

The many different ways of defining a score $s(\xb)$ can be used to further categorize OOD detection methods.
Unsupervised methods often derive the score 
from the \emph{likelihood} assigned to the input \cite{bishop_novelty_1994,choi_waic_2019,kirichenko_why_2020,ren_likelihood_2019,serra_input_2020,xiao_likelihood_2020,morningstar_density_2021,nalisnick_detecting_2019,bergamin_modelagnostic_2022,maaloe_biva_2019,havtorn_hierarchical_2021}, 
from a \emph{reconstruction} of the input $\xb$ \cite{sakurada_anomaly_2014,xia_learning_2015,lyudchik_outlier_2016,zhou_anomaly_2017,chen_outlier_2017,schlegl_unsupervised_2017,zong_deep_2018,li_madgan_2019,graham_denoising_2023,liu_unsupervised_2023a}, or 
from a hidden \emph{representation} of the input \cite{denouden_improving_2018,hendrycks_using_2019,ahmadian_likelihoodfree_2021,bergman_classificationbased_2020,tack_csi_2020,sehwag_ssd_2021,xiao_we_2021}.  % https://arxiv.org/pdf/2302.10326.pdf
Supervised methods often derive the score 
from the \emph{probabilities} given by the predictive distribution $p(y|\xb)$ \cite{hendrycks_baseline_2017,hendrycks_scaling_2022} or 
from the \emph{logits} of $p(y|\xb)$ \cite{hendrycks_scaling_2022,liu_energy-based_2020}. 
Other methods use a latent \emph{representation} of the input \cite{lee_simple_2018,li_anomaly_2019,ndiour_outofdistribution_2020,cook_outlier_2020,zaeemzadeh_outofdistribution_2021}, similar to unsupervised methods. 
In the following sections, we provide a more in-depth overview of the different approaches to OOD detection following the above categorization into supervised and unsupervised methods. For the supervised methods we further distinguish between whether methods use real OOD data, synthetic OOD data, or no OOD data at all.

@@

% \subsubsection{Model agnostic: Related works}
% Since \textcite{nalisnick_deep_2019} and \textcite{hendrycks_deep_2019}, different test statistics or methodologies for OOD detection using DGMs were proposed. Most of the recent solutions were highly influenced by three major lines of work: \emph{typicality set}, \emph{likelihood ratio} test statistics, and \emph{model misestimation}.

% The typicality set hypothesis was introduced by \textcite{nalisnick_detecting_2019} as a possible explanation for the DGMs assigning higher likelihood to OOD data. The typicality set is the subset of the model full support where the model samples from and this does not intersect with the region of higher likelihood. While the typicality test was introduced for batch-OOD detection, \textcite{morningstar_density_2021} shows that it also works well in the single-sample case. This is also confirmed by our own experiments.

% The likelihood ratio test statistic method by \textcite{ren_likelihood_2019} assumes that every input is composed by a background component and a semantic component. For OOD detection, only the semantic component matters. In addition to a model trained on the in-distribution data, they proposed to train a background model on perturbed inputs data and then for each test example consider as OOD score the likelihood ratio between the two models. \textcite{schirrmeister_understanding_2020}, instead, trained the background model on a more general distribution of images by considering 80 million general tiny images. Similarly to these approaches, \textcite{serra_input_2020} argued that the failure of DGMs is due to the high-influence that the input complexity has on the likelihood. Therefore, they proposed to use a general lossless image compression algorithm as a background model.
% All these methods, however, require additional knowledge of the OOD data for either choosing an image augmentation procedure to perturb the input data or for choosing a specific compressor.

% Another line of works blame the models themselves and not the test statistics. \textcite{zhang_understanding_2021} argued that model misestimation is the main cause of higher likelihood assigned to OOD data. This can be due to both the model architecture and the maximum likelihood objective. \textcite{kirichenko_why_2020} and \textcite{schirrmeister_understanding_2020} showed that normalizing flows can achieve better OOD performance despite achieving a worse likelihood if one changes some model design choices.
% %
% Other works in the literature focused on deriving specific test statistics that work only for a specific model, for example for VAEs \parencite{xiao_likelihood_2020, maaloe_biva_2019, havtorn_hierarchical_2021}, or for normalizing flows \parencite{kirichenko_why_2020, ahmadian_likelihoodfree_2021}.

% %All the solutions we presented can be described in terms of the following two categories: \emph{model-agnostic} vs \emph{model-specific} test statistics and statistics \emph{with OOD assumptions} vs statistics \emph{with no OOD assumptions}. In this work, we are mostly interested in model-agnostic scores with no OOD assumptions, i.e.\@ scores that can be directly computed with any models and that do not need any explicit assumption on the OOD data.

% As mentioned in the introduction, we frame the OOD detection problem in terms of statistical tests problem. Recently, \textcite{haroush_statistical_2021} showed that adopting hypothesis testing at the layer and channel level of a neural network can be used for OOD detection in the discriminative setting. They used both Fisher's method and Simes' method to combine class-conditional $p$-values computed for each convolutional and dense layer of a deep neural network. We focus on the unsupervised setting using DGMs and use hypothesis testing on statistics that can be computed on all differentiable DGM. As already explained in section \cref{sec_modelagnostic:combination_p_values}, \textcite{morningstar_density_2021} considered the combination of different statistics for OOD detection. The main difference with their approach is that we propose statistics that can be applied to any differentiable generative model and combine them by using Fisher's method, which takes advantage of using only one-sided independent statistics. Concurrently, \textcite{choi_robust_2021} derived the score statistic by starting from the likelihood ratio statistic and applying a Laplace approximation. They computed the score statistic only for certain layers of the model and for a specific example, the OOD score is given by the infinity norm of these different layer scores after a ReLU operation. Our procedure differs both in the derivation of the score statistic and its usage since we compute the score statistic for the entire model.

% \subsubsection{Hierarchical: Related works}

% So far, no reliable direct likelihood-based method has been found for fully unsupervised deep generative model OOD detection.
% A major line of work considers developing new scores that are more reliable than the likelihood.
% This includes the \textit{typicality} test presented by \textcite{nalisnick_detecting_2019} which is an OOD detection test based on the typicality of a batch of potentially OOD examples.
% This approach however requires a batch of examples from the same class (OOD or not) which limits its practical applicability.
% In \textcite{ren_likelihood_2019}, the \textit{likelihood ratio} between a primary model and a background model was shown to be an effective score for OOD detection.
% However, to train the background model, the in-distribution data is perturbed via a data augmentation technique that is designed with knowledge about the confounding factors between the in-distribution data and the OOD data. Furthermore, it is tuned towards high performance on a known OOD dataset.
% \textcite{serra_input_2020} take a similar approach and attribute the failure to detect OOD data to the high influence of the input complexity on the likelihood and choose a generic lossless compression algorithm as the background model.
% Although this method gives good results, no single best choice of compression algorithm exists for all types of OOD data, and any particular choice encodes prior knowledge about the data into the detection method.
% Both these methods can be seen as correcting for low-level features of the OOD data being assigned high model likelihood by using a second model focused exclusively on these features.

% Similar to these methods, the majority of the approaches to OOD detection make assumptions about the nature of the OOD data.
% The assumptions encompass using labels on the in-distribution data \cite{hendrycks_baseline_2017, liang_enhancing_2018, alemi_uncertainty_2018, lee_simple_2018, lakshminarayanan_simple_2017}, examples of OOD data \cite{hendrycks_deep_2019}, augmenting in-distribution data to mimic it \cite{ren_likelihood_2019}, or assuming a certain data type \cite{serra_input_2020}.
% Any of these assumptions encode implicit biases into the model about the attributes of OOD data which, in turn, might impair performance on truly unknown data examples (unknown unknowns).

% % Introduce the contrast to results with VAEs
% While some of these methods achieve very good results on OOD detection with autoregressive models \cite{oord_pixel_2016, salimans_pixelcnn_2017} and invertible flow-based models \cite{kingma_glow_2018}, it was recently shown that they can be much less effective for VAEs \cite{xiao_likelihood_2020} highlighting the need for a more reliable OOD score for VAEs.
% Although VAEs have the same failure cases as autoregressive and flow-based models, the caveat is that the difference in the likelihood is generally not as big and reconstructions of OOD can be surprisingly good \cite{xiao_likelihood_2020}.
% \textcite{xiao_likelihood_2020} alleviate this by refitting the inference network, as previously proposed by \textcite{cremer_inference_2018, mattei_refit_2018}, to a potentially OOD example and measuring the so-called \textit{likelihood regret}.
% However, refitting the inference network can be computationally expensive, especially for the large hierarchical VAEs that are used to model complex data \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}. Furthermore, this scales poorly to large amounts of potentially OOD examples as the optimization is done per example.

% A few methods have approached OOD detection in a completely unsupervised fashion \cite{maaloe_biva_2019, choi_waic_2019, xiao_likelihood_2020}.
% The work of \textcite{maaloe_biva_2019} is the most related to ours. They introduce BIVA, a deep hierarchy of stochastic latent variables with a top-down and bottom-up inference model and achieve state-of-the-art likelihood scores. 
% They also provide early results indicative that a looser likelihood bound may have value in OOD detection.
% In this paper, we provide an explanation of those results, and significantly improve upon them.

% Out-of-distribution detection and anomaly detection are both related concepts in machine learning and deep learning, but they address slightly different problems and have distinct focuses.

% Out-of-Distribution Detection (OOD Detection):
% Out-of-distribution detection involves determining whether a given input or data point belongs to the same distribution as the training data. In other words, it helps identify whether an input is coming from a category or distribution that the model has never encountered during training. This is especially important for safety-critical applications where the model should not make predictions on inputs that are significantly different from what it has seen before. OOD detection aims to prevent the model from making confident predictions on unfamiliar inputs that might lead to incorrect results.
% Anomaly Detection:
% Anomaly detection, on the other hand, focuses on identifying rare and unusual instances within the data, regardless of whether they belong to the same distribution as the training data or not. Anomalies can be defined as data points that deviate significantly from the norm or expected behavior. Anomaly detection is used to find instances that are different from the majority and might indicate potential errors, fraud, faults, or other unusual occurrences.
% In summary, the key differences are:

% Focus:
% OOD Detection: Focuses on detecting inputs that are coming from distributions that differ significantly from the training data distribution.
% Anomaly Detection: Focuses on identifying rare and unusual instances, regardless of whether they come from a different distribution or not.
% Purpose:
% OOD Detection: Primarily used for safety and robustness, ensuring that the model doesn't make confident predictions on unfamiliar data.
% Anomaly Detection: Used for various applications such as fraud detection, fault detection, and identifying outliers in the data.
% Detection Approach:
% OOD Detection: Often involves measuring the uncertainty of the model's predictions. High uncertainty indicates that the input might be out-of-distribution.
% Anomaly Detection: Focuses on detecting data points that are statistically rare or far from the norm.
% It's worth noting that while the concepts are distinct, they can sometimes overlap. An out-of-distribution data point could also be considered an anomaly, but not all anomalies are necessarily out-of-distribution. Anomaly detection methods might use OOD detection techniques as part of their process to flag anomalous instances that are not in line with the expected distribution.

% Nice review:
% https://openaccess.thecvf.com/content/CVPR2023W/VAND/papers/Graham_Denoising_Diffusion_Models_for_Out-of-Distribution_Detection_CVPRW_2023_paper.pdf

% \cite{kabir_neural_2018} Survey of uncertainty quantification in deep learning.

% https://arxiv.org/pdf/2306.05674.pdf

@@

>>>>

While the papers in this thesis were written to be self-contained, page limits and the need to maintain a specific focus forced us to limit the scope of the technical background that each paper provides. In this chapter, we present a more in-depth overview of the technical background relevant to the papers in this thesis. 
We start by quantifying the concept of uncertainty relating it to information and probability. We then introduce the task of out-of-distribution detection and review the existing litterature on the topic. 
Then, we describe variational autoencoders, how they can be used to learn representations of speech, and some of their challenges. Finally, we introduce the concept of self-supervised learning and highlight important differences to variational autoencoders.

@@

In society, the concept of uncertainty goes by many names, and its meaning can vary depending on the specific context. However, across most quantitative scientific fields, a consensus definition appears to align with the following \cite{hubbard_how_2014}:
%
\begin{center}
    \textit{Uncertainty is the lack of certainty; a state of limited knowledge where it is impossible to exactly describe the existing state, a future outcome, or more than one possible outcome.}
\end{center}
%
Although such a purely lexical definition of uncertainty might prompt a philosophical inquiry, it highlights an important connection between uncertainty, information and probability. When information is limited, we can describe the state of the world only with uncertainty; we are simply not certain. In that sense, uncertainty is a word we use to refer to missing information, whether we know what is missing or not. 
A natural way to describe an uncertain world state is to use probabilities. To our luck, information theory provides exactly that; a mathematically rigorous method for quantifying information through the language of probability theory. In the following, we shall see that, in this context, uncertainty can be understood as the information entropy of a random variable \cite{mackay_information_2003}.

@@

The information entropy of a discrete random variable $X$ is defined as the \emph{expected information content} of an outcome of $X$,
%
\begin{equation} \label{eq:information-entropy}
    H(X) = \mathbb{E}_X\left[I_X(x)\right] = - \sum_{x\in X} p_X(x) \log p_X(x) \enspace .
\end{equation}
%
This means that $H(X)$ measures the amount of information we can expect to gain from observing an outcome of $X$, when we know only its distribution $p_X(x)$. If all possible outcomes are equally likely, then $H(X)$ is maximized, and if only one outcome is possible, then $H(X)$ is minimized.

@@

To generalize the concept of information entropy to continuous random variables, Shannon originally replaced the sum over the probability mass function in \eqref{eq:information-entropy} with an integral over the probability density function, as suggested by the definition $H(X) = \mathbb{E}_X\left[I_X(x)\right]$. This approach leads to the definition of the information entropy of a continuous random variable $X$ as, 
%
\begin{equation} \label{eq:differential-entropy}
    H(X) = - \int_{\mathcal{X}} p_X(x) \log p_X(x) \, dx \enspace ,
\end{equation}
%
which is called the \emph{differential entropy} \cite{shannon_mathematical_1948}. Here, $\mathcal{X}$ is the support of $X$.

However, the differential entropy does not have several of the desirable properties of the discrete version; it can be negative, it is not invariant under a change of variables, and it is not dimensionally correct%
\footnote{\label{fn:dimensional-analysis-of-differential-entropy}
    Information entropy has the same dimensionality as information content which is typically measured in units of bits, but depends on the base of the logarithm. 
    However, the bit is not a base unit of the International System of Units (SI), nor is it an official derived unit.
    A bit is quite simply a number, specifically 0 or 1. If something is said to have a size of 4 bits, it means that it can be described with 4 binary digits, i.e. 4 numbers, each either 0 or 1.

@@

\section{Out-of-distribution detection} \label{sec:out-of-distribution-detection}
% 
% OUTLINE: C.f. https://openaccess.thecvf.com/content/CVPR2023W/VAND/papers/Graham_Denoising_Diffusion_Models_for_Out-of-Distribution_Detection_CVPRW_2023_paper.pdf
% - Formalize the problem. https://arxiv.org/pdf/2302.10326.pdf
%   - Assume we have a dataset $\mathcal{D}_{\text{train}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{train}}$ or not. Given a test point $\xb$, the model outputs a score $s(\xb)$ that indicates whether $\xb$ is in-distribution or out-of-distribution. 
%   - In the context of neural networks, this problem dates back to at least 1994 \cite{bishop_novelty_1994}. 
%
% - Categorisation into supervised and unsupervised. 
%   - Supervised: Whether p(y|x) should be trusted for a given x. 
%   - Unsupervised: Whether x should be trusted and used at all. 
%
In this section we will introduce out-of-distribution (OOD) detection and relate it to the concept of uncertainty. 
To supplement the short related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic}, we will then provide a concise review of the literature on out-of-distribution detection.

\subsection{An overview of the task}
%
OOD detection is the task of identifying test data that is unlikely to originate from the distribution of the training data and, in the context of neural networks, dates back several decades \cite{bishop_novelty_1994, chang_figure_1993}.
In the general case, we assume that we have a domain of in-distribution data, $\mathcal{D}_{\text{in}}$, and we would like to build a model that can be used to assess whether a test data point $\xb$ originates from that domain or not. 
Since such OOD data is unknown to the model, it is, by definition, a source of epistemic uncertainty. This makes OOD detection closely related to uncertainty quantification. 
% Only in cases where strong prior knowledge of the data is incorporated into a model, can we expect it to be able to make reliable predictions on OOD data.

% In the literature, the task is also referred to as open set recognition \cite{bendale_open_2016, cardoso_weightless_2017, panaredabusto_open_2017}, outlier detection \cite{xu_deep_2014, mor_confidence_2018}, or selective prediction \cite{geifman_selective_2017} . 
OOD detection bears many similarities with anomaly detection, novelty detection, open set recognition, and outlier detection \cite{yang_generalized_2022}. 
Some of these differences are subtle, and the terminology is not always used consistently in the literature. To provide some clarity, we will briefly outline the taxonomy of \textcite{yang_generalized_2022}: 
Outlier detection most clearly differs from the other tasks by directly processesing all observations and aiming to select outliers from a single contaminated dataset. 
The remaining tasks differ in whether they detect both covariate and semantic shift and require the simultaneous classification of in-distribution classes. %, and whether the in-distribution data contains more than one class. 
Anomaly detection deals with multiclass data and detects both covariate and semantic shift without requiring simultaneous classification of in-distribution classes whereas novelty detection, open set recognition and OOD detection are usually concerned with semantic shifts. 
Although a vague difference, novelty detection usually defined in terms of a class of normal data, while OOD detection centers around the distribution of the training data. Different from novelty detection, OOD detection methods sometimes draw on examples of OOD data or require the simultaneous classification of in-distribution classes, although this is not always the case. 
OOD detection benchmarks almost always take OOD data to be from external datasets different from the training dataset which distinguishes it from open set recognition that usually uses a single dataset split into ID and OOD data. 
% Another closely related field is that of network-based uncertainty estimation \cite{neal_bayesian_1995, hernandez-lobato_probabilistic_2015, gawlikowski_survey_2023}. %[14, 10, 36, 20]. 
Although we share the sentiment of \textcite{yang_generalized_2022}, who propose to unify the tasks as ``generalized out-of-distribution detection'', this thesis will follow the nomenclature used in the works most related to ours and refer to the task as out-of-distribution detection.

Recently, several approaches for deep neural networks have been developed that formally address the rejection of samples $\xb\notin\mathcal{D}_{\text{in}}$. 
One way to categorize the different approaches is based on whether the underlying model is a classifier that estimates a conditional probability distribution $p(y|\xb)$ over some target variable $y$, or a model that learns a probability density $p(\xb)$ over the input itself. We might refer to the former as \emph{supervised} OOD detection, and the latter as \emph{unsupervised} OOD detection \cite{graham_denoising_2023,liu_unsupervised_2023a}. It is important to note that this distinction relates only to the availability of some target value, $y$ - not whether OOD data is available for supervision\footnote{
    In our overview, we distinguish between supervised and unsupervised out-of-distribution by letting methods be classified as supervised that use any kind of target value $y$, whether it relates to the original task, available OOD data, or both. This is the same distinction made by \textcite{graham_denoising_2023} and \textcite{liu_unsupervised_2023a}. 
    However, it is important to note that in other works, the distinction is made based on whether the model is trained on OOD data or not \cite{hendrycks_baseline_2017,liu_energy-based_2020}. This difference in nomenclature is currently unresolved in the literature. 
}. 
Supervised OOD detection tries to assess whether the model's prediction $\hat{y}$ via $p(y|\xb)$ should be trusted for a given input $\xb$, whereas unsupervised OOD detection judges whether the input $\xb$ should be trusted and used at all.

Central to any OOD detection method is the ability to assign a score $s(\xb) \in \mathbb{R}$ to a given input $\xb$ that indicates the degree to which the input is likely to be OOD. After defining a score, we typically use a validation set to tune a threshold $\tau$ such that $\xb$ is considered OOD if $s(\xb) > \tau$. The threshold is typically chosen such that performance on the validation set is above some level, for instance by imposing constraints on the recall and precision. Many works also evaluate the performance of OOD detection methods using the area under the receiver operating characteristic (AUROC) curve which does not require a threshold to be chosen.

The many different ways of defining a score $s(\xb)$ can be used to further categorize OOD detection methods.
Unsupervised methods often derive the score 
from the \emph{likelihood} assigned to the input \cite{bishop_novelty_1994,choi_waic_2019,kirichenko_why_2020,ren_likelihood_2019,serra_input_2020,xiao_likelihood_2020,morningstar_density_2021,nalisnick_detecting_2019,bergamin_modelagnostic_2022,maaloe_biva_2019,havtorn_hierarchical_2021}, 
from a \emph{reconstruction} of the input \cite{sakurada_anomaly_2014,xia_learning_2015,lyudchik_outlier_2016,zhou_anomaly_2017,chen_outlier_2017,schlegl_unsupervised_2017,zong_deep_2018,li_madgan_2019,graham_denoising_2023,liu_unsupervised_2023a}, or 
from a hidden \emph{representation} of the input \cite{denouden_improving_2018,hendrycks_using_2019,ahmadian_likelihoodfree_2021,bergman_classificationbased_2020,tack_csi_2020,sehwag_ssd_2021,xiao_we_2021}.  % https://arxiv.org/pdf/2302.10326.pdf
Supervised methods often derive the score 
from the \emph{probabilities} given by the predictive distribution $p(y|\xb)$ \cite{hendrycks_baseline_2017,hendrycks_scaling_2022}, 
from the \emph{logits} of $p(y|\xb)$ \cite{hendrycks_scaling_2022,liu_energy-based_2020}, or 
use a latent \emph{representation} of the input \cite{lee_simple_2018,li_anomaly_2019,ndiour_outofdistribution_2020,cook_outlier_2020,zaeemzadeh_outofdistribution_2021}, similar to unsupervised methods. 
In the following sections, we provide a more in-depth overview of the different approaches to OOD detection following the above categorization into supervised and unsupervised methods. For the supervised methods we further distinguish between whether methods use real OOD data, synthetic OOD data, or no OOD data at all.

@@

@@

A variational autoencoder (VAE) \cite{kingma_autoencoding_2014,rezende_stochastic_2014} is a type of deep generative model and an instance of a latent variable model. In the following, we will introduce VAEs starting from the general definition of latent variable models. We then derive the maximum likelihood objective used to train VAEs, discuss its different components, and highlight some central challenges. Finally, we discuss some of the most important extensions to VAEs made model speech data.

The fundamental assumption underlying the definition of latent variable models is that a data point $\xb$ is created via a generative process that involves one or more unobserved, stochastic latent variables  $\zb$. 
Latent variable model design centers around capturing this generative process and learning to generate new data points $\xb$. The generative process is usually defined by a joint distribution $p(\xb, \zb)$ over the data and latent variables. In many models, the joint distribution is factorized as $p(\xb,\zb) = p(\xb|\zb)p(\zb)$, including Gaussian mixture models \cite{dempster_maximum_1977}, hidden Markov models \cite{rabiner_tutorial_1989}, and probabilstic principal component analysis \cite{tipping_probabilistic_1999}. In this case, we can write the marginal distribution of the data as,
%
\begin{equation} \label{eq:latent-variable-model-marginal-distribution}
    p(\xb) = \int p(\xb|\zb)p(\zb) \, d\zb \enspace .
\end{equation}
%
The ability to infer the latent variables $\zb$ from a given data point $\xb$ is of interest in applications that focus on representation learning. Inference is usually performed by computing the posterior distribution $p(\zb|\xb)$, which can be computed using Bayes' rule,
%
\begin{equation} \label{eq:latent-variable-model-posterior-distribution}
    p(\zb|\xb) = \frac{p(\xb|\zb)p(\zb)}{p(\xb)} \enspace .
\end{equation}
%
The marginal distribution $p(\xb)$ involves an integral over all possible values of the latent variable $\zb$ that is in many cases analytically intractable. This is especially true when the latent variable space is high-dimensional or if the conditional distribution $p(\xb|\zb)$ is complex, as with neural networks or non-conjugate models. Approximate methods such as variational inference \cite{jordan_introduction_1999} and Markov Chain Monte Carlo (MCMC) \cite{mohamed_monte_2019} are therefore often used to compute the marginal and posterior distributions but these methods can be computationally expensive.

When the data is high-dimensional, an obvious choice for modelling the generative process is to use a deep neural network. This makes the marginal and posterior distributions intractable and requires approximate methods for inference. Variational autoencoders \cite{kingma_autoencoding_2014,rezende_stochastic_2014} use variational inference to learn a latent representation of the input data. 
Variational inference introduces an approximate posterior distribution $q(\zb|\xb)$ that is optimized to be close to the true posterior distribution $p(\zb|\xb)$. 
We can write the marginal distribution of the data as,
%
\begin{align}
    a
\end{align}
%

VAEs amortize the cost of inference by parameterizing the approximate posterior with a neural network $q_\phib(\zb|\xb)$ that can produce approximate posterior distributions for any input $\xb$.

trained jointly with the generative model $p_\thetab(\xb|\zb)$ by maximizing the evidence lower bound (ELBO),

Any VAE aims to model this generative process by learning a generative model $p_\thetab(\xb, \zb)$

For a more background on VAEs and a comparison with self-supervised methods, we refer the reader to our paper \textcite{borgholt_brief_2022} included in \cref{app:paper-brief}.
